A Theory for Database Transactions with no Database

### 2019-03-05 2018 Transactions/articleTransactions.tex
minor improvements

@@

	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
\end{array}	
\label{eqn:def occurs}
\end{equation}
	This definition expands the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful history structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	To describe the growth of a stateful history structure we can define stateful history structures recursively:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	Information systems can register changes (as opposed to states) to avoid duplicating most of a previous state into a new state.
	Such changes are called {\em transactions}.
	For the purpose of computing a new state ($s_2$) from a given state ($s_1$),
	we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	In a stateful history structure we can represent transactions by means of a function $\delta$,
	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
\begin{equation}
	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
\label{def:delta}
\end{equation}
	The transaction caused by event $e$ can be described by:
\begin{eqnarray}
&&	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
		\id{state}(e)\ =\ s&\wedge\\
		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
	\end{array}
\label{req:transaction}
\end{eqnarray}
	We can define a stateful history structure,
	in which the state after every event is computed from a change, $\Delta$.
	For this purpose we must designate one specific state, $\mathcal I\in\mathbb S$, as the initial state.
	We must additionally specify an existing event $e_1$ in order to compute $\id{state}(e_1)\oplus\Delta$.
	The following mapping $\id{occurByDelta}_{P,e_1,e,\Delta}$ defines how an SHS grows by one event ($e$)
	using only the change between the state of event $e_1$ to the new event $e$:
\begin{eqnarray}
&&\begin{array}{l}
	\id{occurByDelta}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
				&=&\mathcal I\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}
		\end{array}
\end{array}
\label{req:SHS by Delta}
\end{eqnarray}
	This mapping serves as recursion step in the following definition of a stateful history structure
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def SHS by Delta}
\end{eqnarray}

	At this point, let us reflect on storage.
	If we store a stateful history structure entirely,
	one state is added to the storage for every event in $\Events$.
	However, if changes are on average much smaller than states, it can be more economical to store only the changes.
	The following mapping transforms a stateful history structure to an incremental history structure:
\begin{eqnarray}
&&\begin{array}{l}
	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l}
		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
		\end{array}
\end{array}
\label{req:shs2ihs}
\end{eqnarray}
	If we want to build an IHS event-by-event, we need an incremental definition.
	For this purpose we define an {\em incremental history structure} ($\id{IHS}$)
	as a history structure that ``stores'' changes instead of states.
	The following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
	adds one event ($e$) to an IHS:
\begin{eqnarray}
\begin{array}{l}
	\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra\\
	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
		\text{Let }\delta'(x,y)&=&\Delta\text{, }&\text{if }x=e_1\wedge y=e\wedge x\in P\\
				       &=&(\id{state}(e_1)\oplus\Delta)\ominus\id{state}(x)\text{, }&\text{if }x\neq e_1\wedge y=e\wedge x\in P\\
				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}\\
		\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge e_1\notin P\\
				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge e_1\notin P\\
			\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra}
		\end{array}
\end{array}
\label{req:increment}
\end{eqnarray}
	This mapping can be used as recursive step in the following (recursive) definition of predicate $\id{IHS}$, which characterizes an incremental history structure:
\begin{eqnarray}
	&&\id{IHS}\la \emptyset, \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{IHS}\la \Events, \id{prec}, \delta, \iota\ra\\
	\ \ \Rightarrow\ \id{IHS}(\id{increment}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \delta, \iota\ra)
\end{array}
\label{req:recursive def IHS}
\end{eqnarray}

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} means that a state change caused by a sequence of events can be reproduced as a single event.

>>>>

	An interesting read is {\tt\small https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
	It explains how modern distributed database cope with ACID properties in a was that makes sense.

@@

	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
\end{array}	
\label{eqn:def occurs}
\end{equation}
	This definition expands the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful history structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	To describe the growth of a stateful history structure by the event,
	we define $\id{SHS}$ recursively using $\id{occurs}_{P,e,s}$ as recurring step:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	Information systems can register changes (as opposed to states) to avoid duplicating most of a previous state into a new state.
	Such changes are called {\em transactions}.
	For the purpose of computing a new state ($s_2$) from a given state ($s_1$),
	we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	In a stateful history structure we can represent transactions by means of a function $\delta$,
	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
\begin{equation}
	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
\label{def:delta}
\end{equation}
	A transaction caused by event $e$ can be described by:
\begin{eqnarray}
&&	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
		\id{state}(e)\ =\ s&\wedge\\
		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
	\end{array}
\label{req:transaction}
\end{eqnarray}
	We can define a stateful history structure,
	in which the state after every event is computed from a change, $\Delta$.
	For this purpose we must designate one specific state, $\mathcal I\in\mathbb S$, as the initial state.
	We must additionally specify an existing event $e_1$ in order to compute $\id{state}(e_1)\oplus\Delta$.
	The following mapping $\id{occurByDelta}_{P,e_1,e,\Delta}$ defines how an SHS grows by one event ($e$)
	using only the change between the state of event $e_1$ to the new event $e$:
\begin{eqnarray}
&&\begin{array}{l}
	\id{occurByDelta}_{P,e_1,e,\Delta}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
				&=&\mathcal I\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}
		\end{array}
\end{array}
\label{req:SHS by Delta}
\end{eqnarray}
	This mapping serves as recursion step in the following definition of a stateful history structure
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def SHS by Delta}
\end{eqnarray}

	At this point, let us reflect on storage.
	If we store a stateful history structure entirely,
	one state is added to the total storage requirement for every event in $\Events$.
	However, if the average change is typically smaller than the average state, it can be more economical to store only the changes.
	For this purpose we define an incremental history structure (IHS)
	as a history structure that ``stores'' changes instead of states.
	A {\em incremental history structure} is a quadruple $\la\Events, \id{prec}, \delta, \iota\ra$ in which
	$\la \Events, \id{prec}\ra$ is a history structure,
	$\delta(e_1,e_2)$ represents the change $\id{state}(e_2)\ominus\id{state}(e_1)$ for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta}),
	and $\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
	Since histories are built event-by-event, we give an incremental definition.
	The following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
	adds one event ($e$) to an IHS:
\begin{eqnarray}
\begin{array}{l}
	\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra\\
	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}
		\text{Let }\delta'(x,y)&=&\Delta\text{, }&\text{if }x=e_1\wedge y=e\wedge x\in P\\
				       &=&(\id{state}(e_1)\oplus\Delta)\ominus\id{state}(x)\text{, }&\text{if }x\neq e_1\wedge y=e\wedge x\in P\\
				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}\\
		\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge e_1\notin P\\
				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge e_1\notin P\\
			\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra}
		\end{array}
\end{array}
\label{req:increment}
\end{eqnarray}
	This definition inserts a new event, $e$, into a history structure with $e_1$ as its predecessor.
	The new state, $\id{state}(e)$, is computed by $\id{state}(e_1)\oplus\Delta$; let us call it $s$.
	Beside $e_1$, there may be other events preceding $e$ in the predecessor set $P$.
	The change from every predecessor $x$ can be computed by $s\ominus\id{state}(x)$.
	All these changes are added to the function $\delta$.
	The function $\iota$ is there for initialization purposes.
	To get an initial history structure (with no events in it) going,
	we must add events while there are no other events available as predecessor.
	We use the empty predecessor set as condition to signal that there is no previous state.
	The mapping $\id{increment}_{P,e_1,e,\Delta}$ can be used as recursive step in defining predicate $\id{IHS}$,
	which characterizes an incremental history structure:
\begin{eqnarray}
	&&\id{IHS}\la \emptyset, \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{IHS}\la \Events, \id{prec}, \delta, \iota\ra\\
	\ \ \Rightarrow\ \id{IHS}(\id{increment}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \delta, \iota\ra)
\end{array}
\label{req:recursive def IHS}
\end{eqnarray}
The following mapping transforms a stateful history structure to an incremental history structure:
\begin{eqnarray}
&&\begin{array}{l}
	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l}
		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
		\end{array}
\end{array}
\label{req:shs2ihs}
\end{eqnarray}
	Vice versa, we can also map an incremental history structure to a stateful history structure:
\begin{eqnarray}
&&\begin{array}{l}
	\id{ihs2shs}(\la\Events, \id{prec}, \delta, \iota\ra)\\
	\ \ =\ \begin{array}[t]{@{}l}
		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\\
		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
		\text{Let }\forall e\in\id{initial}:\ \id{state}(e)=\mathcal I\oplus\iota(e)\\
		\text{in }\la\Events, \id{prec}, \id{state}\ra
		\end{array}
\end{array}
\label{req:ihs2shs}
\end{eqnarray}
	The\marginpar{claim} fact that $\id{shs2ihs}$ and $\id{ihs2shs}$ are each other's inverse proves that the choice to represent
	history by an IHS or SHS can be made on practical grounds.

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} guarantees all-or-nothing outcomes.
	It means that the overall state change caused by a sequence of events can be committed as a whole or rejected completely.
